A Stopped Negative Binomial Distribution

### 2016-03-01 stat-and-prob-letters/snb-paper.tex
Draft for my review.

The rest of this paper derives the distribution for the number of
patients needed to end a trial and explores the characteristics and
properties of this distribution.
The next section introduces our notation and basic results
including the density of the distribution along with a description of
it's relation to other distributions. Sections 2 derives the distribution
based on a defined Bernoulli process and give some basic properties.
Section 3 provides a connection to the Binomail tail probability.
Section 4 derives the posterior distribution using a Beta prior.
Section 5 provides a brief discussion on the use of the distribution
in clinical trials along with future avenues for generalization.

@@

\begin{prop}
The distribution of the stopping time
$\argminl_k \left\{ X_k \geq s \cup X_k \leq k-t \right\}$
is equal to Eqn.~\ref{eqn:pmf}.
\end{prop}
\begin{proof}
%The proof will proceed in two parts. First, a combinatorial justification 
%will be given for the probability mass value on each element of the support. 
%Second, it will be shown that the sum of the masses over the support sums to 
%one.

@@

\section{Connection to the Binomial Tail Probability}

\begin{figure}[t!]
\centering
\subfloat[The SNB distribution.]{\includegraphics[width=0.5\textwidth]{snb_density.pdf}}
\hfill
\subfloat[The Binomial distribution.]{\includegraphics[width=0.5\textwidth]{bin_density.pdf}}
\caption{
SNB(0.2, 2, 12) with mass contributed from 
$s$ successes (red) or $t$ failures (teal) along with Bin(0.2, 13) with
at least 2 successes (red) or fewer (teal).
}
\label{fig:kane_viz}
\end{figure}

\begin{prop} Let $Y$ be distributed as SNB($p$, $s$, $t$) and let 
$B$ be distributed Binomial with size $n=s+t-1$ and success probability
$p$. Then
\begin{equation}
\mathbb{P}[B \geq s] = \mathbb{P} [Y \leq n | \text{\#success} = s].
\end{equation}
That is, the probability that the number of successes is at least $s$
in the Binomial model is the same that the trial stops with $s$ 
successes in the SNB model.
\end{prop}
\begin{proof}
The Binomial tail probability is
\begin{align*}
\mathbb{P}[B \geq s] &= \sum_{k=s}^{s+t-1} {n \choose k} p^k (1-p)^{n-k} \\
  &= 1 - \sum_{k=0}^{s-1} p^k (1-p)^{n-k} \\
  & 1 - \mathcal{I}_{1-p}.
\end{align*}
Next, consider the SNB probability
\begin{equation*}
\mathbb{P} [Y \leq n | \text{\#success} = s] = \sum_{k=s}^{s+t-1}
  {k-1 \choose s-1} p^s (1-p)^k-s.
\end{equation*}
Let $i=k-s$ then using the fact that ${i+s-1 \choose s-1} = {i+s-1 \choose i}$
the summation can be rewritten as
\begin{align}
\mathbb{P} [Y \leq n | \text{\#success} = s] &= \sum_{i=0}^{t-1} 
  {i+s-1 \choose i} p^s (1-p)^i\\
  &= 1 - \mathcal{I}_{1-p}(t, s).
\end{align}
\end{proof}

@@

\begin{prop} Let $S$ be distributed SNB with parameters $p$, $s$, and $t$.
Then the moment generating function (MGF) of $S$ is
\begin{equation} \label{eqn:mgf}
\mathbb{E}~e^{xS} = \frac{p e^x}{1 - qe^x} \mathcal{I}_{qe^x} (s, t) + 
  \frac{qe^x}{1-pe^x} \mathcal{I}_{pe^x}(s, t)
\end{equation}
for $q = 1-p$ when $x \leq -\log(q)$ and $x \leq -\log(p)$.
\end{prop}
\begin{proof}
By definition, the MGF of the SNB is:
\begin{equation*}
\mathbb{E}~e^{xS} = \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s q^{k-s} e^{kx} 
  + \sum_{k=t}^{s+t-1} {k-1 \choose k-t} p^{k-t} q^t e^{kx}
\end{equation*}
and can be rewritten as:
\begin{equation} \label{eqn:first_sum}
\mathbb{E}~e^{xS} = \sum_{k=s}^{s+t-1}{k-1 \choose s-1} (pe)^{sx} (qe^x)^{k-s} 
  + \sum_{k=t}^{s+t-1}{k-1 \choose k-t} (qe^x)^t (pe^x)^{k-t}.
\end{equation}
Taking the first summation in Equation \ref{eqn:first_sum}:
\begin{align*}
\sum_{k=s}^{s+t-1}{k-1 \choose s-1} (pe)^{sx} (qe^x)^{k-s} &= 
  \left(\frac{pe^x}{1 - qe^x}\right)^s \ \ \sum_{k=s}^{s+t-1} {k-1 \choose s-1} 
    (qe^x)^{k-s} (1-qe^x)^s) \\
  &= \left(\frac{pe^x}{1 - qe^x}\right)^s \mathcal{I}_{qe^x}(s, t).
\end{align*}
Since the incomplete beta function has support on zero to one
$qe^x \leq 1$. This implies that $x \leq -\log(q)$.

@@

Let us assume that $p$ has a Beta distribution, with constant prior
parameters $\alpha$ and $\beta$. Then the closed form
posterior distribution of the SNB is
\begin{align}
f \left(k | \mathbf{b}, \alpha, \beta \right) &= 
  {k-1 \choose s-1} \frac{B\left(\alpha+s, k-s+\beta \right)}{B(\alpha, \beta)} 
  \ I_{\{s \leq k \leq s+k-1\}} + \nonumber \\
  & {k-1 \choose k-t} 
  \frac{B\left(\alpha + k - t, t+\beta\right)}{B(\alpha, \beta)} 
  \ I_{\{t \leq k \leq s+k-1\}}
\end{align}
where $B$ is the Beta function and $\mathbf{b}$ is the realization
of Bernoulli data.

>>>>

\begin{table}[t!]
\caption{Stopped Negative Binomial Distribution Characteristics}
\label{tab:snb}
\begin{center}
\begin{tabular}{|c|l|} \hline
Notation & SNB($p$, $s$, $t$) \\ \hline
Parameters & $p$ the success probability ($q = 1-p$) \\
           & $s$ the number of successes before stopping \\
           & $t$ the number of failures before stopping \\ \hline
Support & min$(s,t) \leq k \leq s+t-1$  \\ \hline
PMF & ${k-1 \choose s-1} p^s (1-p)^{k-s} + {k-1 \choose t-1} (1-p)^s p^{k-t}$\\ \hline
CDF & $2 - \mathcal{I}_{1-p}(k+1, s) - \mathcal{I}_{p}(k+1, t)$\\ 
    & where $\mathcal{I}$ is the regularized incomplete beta function.\\ \hline
Mean & $\frac{s}{p} \mathcal{I}_p(s,t) + \frac{q^{t-1} p^{s-1}}{B(s,t)} +
  \frac{t}{q} \mathcal{I}_q(t,s) + \frac{q^{s-1}p^{t-1}}{B(s,t)}$\\ 
  & where $B$ is the beta function \\ \hline
MGF & $\left(\frac{p e^x}{1 - qe^x}\right)^s 
  \mathcal{I}_{1-qe^x} (s, t) + \left(\frac{qe^x}{1-pe^x}\right)^t 
  \mathcal{I}_{1-pe^x}(t, s) $\\ \hline
\end{tabular}
\end{center}
\end{table}

Tab.~\ref{tab:snb} summarizes the distribution and 
some of its characteristics.
The rest of this paper derives these results
and explores properties of the distribution.
The next section introduces our notation and basic results
including the density of the distribution along with a description of
it's relation to other distributions. Sections 2 derives the distribution
based on a defined Bernoulli process and give some basic properties.
Section 3 provides a connection to the Binomial tail probability.
Section 4 derives the posterior distribution using a Beta prior.
Section 5 provides a brief discussion on the use of the distribution
in clinical trials along with future avenues for generalization.

@@

\begin{prop}
The distribution of the stopping time
$\argminl_k \left[X_k \geq s \cup X_k \leq k-t \right]$
is equal to Eqn.~\ref{eqn:pmf}.
\end{prop}
\begin{proof}
%The proof will proceed in two parts. First, a combinatorial justification 
%will be given for the probability mass value on each element of the support. 
%Second, it will be shown that the sum of the masses over the support sums to 
%one.

@@

\section{Connection Between $S$ and the Binomial Tail Probability}

\begin{prop} \label{binomial_tail}
Let $Y$ be distributed as SNB($p$, $s$, $t$) and let 
$B$ be distributed Binomial with size $n=s+t-1$ and success probability
$p$. Then
\begin{equation}
\mathbb{P}[B \geq s] = \mathbb{P} [Y \leq n\ |\ \text{\#success} = s].
\end{equation}
That is, the probability that the number of successes is at least $s$
in the Binomial model is the same that the trial stops with $s$ 
successes in the SNB model.
\end{prop}
\begin{proof}
The Binomial tail probability is
\begin{align*}
\mathbb{P}[B \geq s] &= \sum_{k=s}^{s+t-1} {n \choose k} p^k (1-p)^{n-k} \\
  &= 1 - \sum_{k=0}^{s-1} {n \choose k} p^k (1-p)^{n-k} \\
  & 1 - \mathcal{I}_{1-p}.
\end{align*}
Next, consider the SNB probability
\begin{align*}
\mathbb{P} [Y \leq n\ |\ \text{\#success}=s] &= \sum_{k=s}^{s+t-1} S(k, p, s)\\
 &= {k-1 \choose s-1} p^s (1-p)^{k-s}.
\end{align*}
Let $i=k-s$ then using the fact that ${i+s-1 \choose s-1} = {i+s-1 \choose i}$
the summation can be rewritten as
\begin{align}
\mathbb{P} [Y \leq n\ |\ \text{\#success} = s] &= \sum_{i=0}^{t-1} 
  {i+s-1 \choose i} p^s (1-p)^i\\
  &= 1 - \mathcal{I}_{1-p}(t, s).
\end{align}
\end{proof}

This result is visualized in Fig.~\ref{fig:snb_bin_compare} for the case
where $s=2$, $t=12$, and $p=0.2$. The probability masses represented in
red are equal as are the masses in teal. The probability that $s$
successes are reached in the SNB process is the same as the binomial 
probability of at least two successes. Likewise, the probability that $t$ 
failures are reached in the SNB process is the same as the binomial
probability of zero or one successes.

\begin{figure}[t!]
\centering
\subfloat[The SNB distribution.]{\includegraphics[width=0.5\textwidth]{snb_density.pdf}}
\hfill
\subfloat[The Binomial distribution.]{\includegraphics[width=0.5\textwidth]{bin_density.pdf}}
\caption{
SNB(0.2, 2, 12) with mass contributed from 
$s$ successes (red) or $t$ failures (teal) along with Bin(0.2, 13) with
at least 2 successes (red) or fewer (teal).
}
\label{fig:snb_bin_compare}
\end{figure}

@@

\begin{prop} Let $Y$ be distributed SNB with parameters $p$, $s$, and $t$.
Then the moment generating function (MGF) of $Y$ is
\begin{equation} \label{eqn:mgf}
\mathbb{E}~e^{xY} = \left(\frac{p e^x}{1 - qe^x}\right)^s 
  \mathcal{I}_{1-qe^x} (s, t) + \left(\frac{qe^x}{1-pe^x}\right)^t 
  \mathcal{I}_{1-pe^x}(t, s)
\end{equation}
for $q = 1-p$ when $x \leq -\log(q)$ and $x \leq -\log(p)$.
\end{prop}
\begin{proof}
By definition, the MGF of the SNB is:
\begin{equation*}
\mathbb{E}~e^{xY} = \sum_{k=s}^{s+t-1} {k-1 \choose k-s} p^s q^{k-s} e^{kx} 
  + \sum_{k=t}^{s+t-1} {k-1 \choose k-t} p^{k-t} q^t e^{kx}
\end{equation*}
and can be rewritten as:
\begin{equation} \label{eqn:first_sum}
\mathbb{E}~e^{xY} = \sum_{k=s}^{s+t-1}{k-1 \choose k-s} (pe)^{sx} (qe^x)^{k-s} 
  + \sum_{k=t}^{s+t-1}{k-1 \choose k-t} (qe^x)^t (pe^x)^{k-t}.
\end{equation}
Taking the first summation in Equation \ref{eqn:first_sum}:
\begin{align*}
\sum_{k=s}^{s+t-1}{k-1 \choose k-s} (pe)^{sx} (qe^x)^{k-s} &= 
  \left(\frac{pe^x}{1 - qe^x}\right)^s \ \ \sum_{k=s}^{s+t-1} {k-1 \choose k-s} 
    (qe^x)^{k-s} (1-qe^x)^s \\
  &= \left(\frac{pe^x}{1 - qe^x}\right)^s \mathcal{I}_{qe^x}(s, t).
\end{align*}
Since the incomplete beta function has support on zero to one
$qe^x \leq 1$. This implies that $x \leq -\log(q)$.

@@
